Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-02-02 paper/Paper.Rnw
further progress editing Methods and associated SI sections

%the beta fucundities version
The \emph{beta-fecundities lottery model}
used $B_i = \eta_i F_\beta^{-1} \circ \varphi(b_i)$, where $\varphi$ is the CDF of the standard normal
distribution and $F_\beta$ is the CDF of a beta distribution 
with shape and scale parameters $0.5$ (probability density function proportional to $x^{-05}(1-x)^{-0.5}$ on
the unit interval). So the beta-fecundities model had parameters $\eta_1$, $\eta_2$ and $\delta$,
and produced fecundities with $U$-shaped, beta distributions between $0$ and $\eta_1$, for $B_1$, and
$\eta_2$, for $B_2$. The mean fecundity for species $i$ was $\eta_i/2$, and larger $\eta_i$ also 
corresponded to more variable fecundities for species $i$.

@@

\subsection{How results were plotted for G2}

@@

%Then a similar para for the beta-fecundity lottery model.
To explore, using the beta fecundities lottery model, the relative magnitudes of the contributions 
to coexistence that come from ATAs versus other mechanisms, we plotted the terms of the decomposition 
(\ref{eq:full_decomp}) for that model against the model parameters $\eta_1$, $\eta_2$, and $\delta$.
%***DAN: finish it
%***DAN: Next steps as of 2023 02 01:
%Fill in the sup mat section for the theory for the beta fecundities lottery model
%fill in the efficient computation section for that same model
%write this section, here, in the main text

>>>>

%the beta fucundities version
The \emph{beta-fecundities lottery model}
used $B_i = \eta_i F_\beta^{-1} \circ \varphi(b_i)$, where $\varphi$ is the CDF of the standard normal
distribution and $F_\beta$ is the CDF of a beta distribution 
with shape and scale parameters $0.5$ (probability density function proportional to $x^{-05}(1-x)^{-0.5}$ on
the unit interval). So the beta-fecundities model had parameters $\eta_1$, $\eta_2$ and $\delta$,
and produced fecundities with $U$-shaped, beta distributions between $0$ and $\eta_1$, for $B_1$, and
$\eta_2$, for $B_2$. The mean fecundity for species $i$ was $\eta_i/2$, and larger $\eta_i$ also 
corresponded to more variable fecundities for species $i$. The parameter $\eta_i$ represents the upper
bound on what fecundities could occur for species $i$.

@@

\subsection{How results were plotted for goal G2}

@@

%Then a similar para for the beta-fecundity lottery model.
To explore, using the beta fecundities lottery model, the relative magnitudes of the contributions 
to coexistence that come from ATAs versus other mechanisms, we plotted the terms of the decomposition 
(\ref{eq:full_decomp}) for that model against the model parameters $\eta_1$, $\eta_2$, and $\delta$,
again labelling regions corresponding to ATA rescue or exclusion.
Some mathematical shortcuts, summarized here, simplified plotting. 
First, we assumed, without loss of generality, that $\eta_1 \leq \eta_2$. To understand coexistence, 
it was then sufficient to decompose GWR of the weaker competitor, species 1. 
Second, we showed that, for the beta fecundities
lottery model, none of the components of (\ref{eq:full_decomp}) depends independently on $\eta_1$ and $\eta_2$;
instead they depend only on $\eta_1/\eta_2$ (SI section \ref{SI-SIsect:TheoryForLottery}). 
So we plotted against $\eta_1/\eta_2$. 
For each combination of $\delta=0.2, 0.4, 0.6$
and for left- and right-tail associated noise, we plotted 
the components of (\ref{eq:full_decomp}) against the upper bound ratio $\eta_1/\eta_2$. 
The simulation methods of \cite{Ellner_2019} were again used.
Computationally efficient means of performing these simulations are in 
SI section \ref{SI-sect:efficient_computation}.

%***DAN: Next steps as of 2023 02 01:
%DONE Fill in the sup mat section for the theory for the beta fecundities lottery model
%fill in the efficient computation section for that same model
%DONE write this section, here, in the main text

### 2023-02-02 paper/SupMat.Rnw
further progress editing Methods and associated SI sections

Recall we took $E_i(t)=B_i(t)$, thereby assuming that fecundity depends strictly on
the environment for the lottery model (both the log-normal fecundities and the beta fecundities versions of the model). 
Recall also that competition, $C_i(t)$, is 
$$C_i(t)=( B_1(t)N_1(t)+B_2(t)N_2(t) )/(\delta N)$$ (again for both versions of the model), which
does not depend on $i$. Species growth rates for the lottery model, $r_i(t)=\ln(N_{i}(t+1)/N_i(t))$, can then be written 
$$r_i(t)=\ln[1-\delta+\delta N B_i(t)/(B_1(t)N_1(t)+B_2(t)N_2(t))]=\ln[1-\delta+E_i(t)/C_i(t)],$$ 
which is an increasing
function of $E_i$ and decreasing function of $C_i$, as was assumed at the outset of Theory.

When $i$ is an invader, the competition experienced by either 
species is obtained from the expression for $C_i=C_j$ by setting $N_i=0$ and $N_j=N$. Thus 
$$C_{i \bs i}(t)=C_{j \bs i}(t)=B_j(t)/\delta=E_j(t)/\delta.$$
Lottery-model values of all the expressions of SI section \ref{SIsect:other_mechanisms} can be obtained 
from the above expression for $C_{i \bs i}$ and the assumption that $E_i(t)=B_i(t)$.
For instance, 
$$\olr_{i \bs i} = \E[r_i(E_i,C_{i \bs i})] = \E[\ln(1-\delta+E_i /C_{i \bs i})] = \E[\ln(1-\delta+\delta B_i/B_j)].$$ 
Likewise, 
$$\olr_{j \bs i} = \E[r_j(E_j,C_{j \bs i})] = \E[\ln(1-\delta+E_j/C_{j \bs i})] = \E[\ln(1-\delta+\delta B_j/B_j)]=0,$$
though we already knew, \emph{a priori}, that this equals $0$. 
An expression for $\olrs_{i \bs i}$ is obtained by replacing $(E_i,C_{i \bs i})=(B_i,B_j/\delta)$
by a random variable with the same marginals, but independent components, $(B_i^\shp,B_j^\shp /\delta)$.
Thus 
$$\olrs_{i \bs i} = \E[\ln(1-\delta+\delta B_i^\shp/B_j^\shp)].$$ 
Likewise, $\olrs_{j \bs i}$ is computed by replacing $(E_j,C_{j \bs i})$ in an analogous way; we can use 
$(E_j^\shp,C_{j \bs i}^\shp)=(B_j^\shp, B_j /\delta)$, so 
$$\olrs_{j \bs i}=\E[\ln(1-\delta+\delta B_j^\shp / B_j)].$$
An expression for $\olrn_{i \bs i}$ is obtained by replacing $(E_i,C_{i \bs i})=(B_i,B_j/\delta)$
by a random variable with the same marginals, but with ATAs removed, $(B_i^\nat,B_j^\nat /\delta)$.
Thus 
$$\olrn_{i \bs i} = \E[\ln(1-\delta+\delta B_i^\nat/B_j^\nat)].$$ 
Likewise, $\olrn_{j \bs i}$ is computed by replacing $(E_j,C_{j \bs i})=(B_j,B_j/\delta)$ in an analogous way.
But note that $B_j$ and $B_j/\delta$ are linearly related for the lottery model, so their association is 
already symmetric, so
so $E_j^\nat=E_j$, $C_{j \bs i}^\nat = C_{j \bs i}$, and $\olrn_{j \bs i}=\olr_{j \bs i}$.
This type of reasoning is applied to the log-normal fecundities and beta fecundities versions of the 
lottery model, specifically, in the subsequent sections.

@@

\subsection{Theory and plotting simplifications for the beta fecundities lottery model}\label{SIsect:TheoryForLotteryBeta}
%***DAN: Fill this in.

>>>>

Recall we took $E_i(t)=B_i(t)$, thereby assuming that fecundity depends strictly on
the environment for the lottery model (both the log-normal fecundities and the beta fecundities versions of the model). 
Recall also that competition, $C_i(t)$, is 
\begin{equation}
C_i(t)=( B_1(t)N_1(t)+B_2(t)N_2(t) )/(\delta N)
\end{equation}
(again for both versions of the model), which
does not depend on $i$. Species growth rates for the lottery model, $r_i(t)=\ln(N_{i}(t+1)/N_i(t))$, can then be written 
\begin{equation}
r_i(t)=\ln[1-\delta+\delta N B_i(t)/(B_1(t)N_1(t)+B_2(t)N_2(t))]=\ln[1-\delta+E_i(t)/C_i(t)],
\end{equation}
which is an increasing
function of $E_i$ and decreasing function of $C_i$, as was assumed at the outset of Theory.

When $i$ is an invader, the competition experienced by either 
species is obtained from the expression for $C_i=C_j$ by setting $N_i=0$ and $N_j=N$. Thus 
\begin{equation}
C_{i \bs i}(t)=C_{j \bs i}(t)=B_j(t)/\delta=E_j(t)/\delta.
\end{equation}
Lottery-model values of all the expressions of SI section \ref{SIsect:other_mechanisms} can be obtained 
from the above expression for $C_{i \bs i}$ and the assumption that $E_i(t)=B_i(t)$.
For instance, 
\begin{equation}
\olr_{i \bs i} = \E[r_i(E_i,C_{i \bs i})] = \E[\ln(1-\delta+E_i /C_{i \bs i})] = \E[\ln(1-\delta+\delta B_i/B_j)].
\end{equation}
Likewise, 
\begin{equation}
\olr_{j \bs i} = \E[r_j(E_j,C_{j \bs i})] = \E[\ln(1-\delta+E_j/C_{j \bs i})] = \E[\ln(1-\delta+\delta B_j/B_j)]=0,
\end{equation}
though we already knew, \emph{a priori}, that this equals $0$. 
An expression for $\olrs_{i \bs i}$ is obtained by replacing $(E_i,C_{i \bs i})=(B_i,B_j/\delta)$
by a random variable with the same marginals, but independent components, $(B_i^\shp,B_j^\shp /\delta)$.
Thus 
\begin{equation}
\olrs_{i \bs i} = \E[\ln(1-\delta+\delta B_i^\shp/B_j^\shp)].
\end{equation}
Likewise, $\olrs_{j \bs i}$ is computed by replacing $(E_j,C_{j \bs i})$ in an analogous way; we can use 
$(E_j^\shp,C_{j \bs i}^\shp)=(B_j^\shp, B_j /\delta)$, so 
\begin{equation}
\olrs_{j \bs i}=\E[\ln(1-\delta+\delta B_j^\shp / B_j)].
\end{equation}
An expression for $\olrn_{i \bs i}$ is obtained by replacing $(E_i,C_{i \bs i})=(B_i,B_j/\delta)$
by a random variable with the same marginals, but with ATAs removed, $(B_i^\nat,B_j^\nat /\delta)$.
Thus 
\begin{equation}
\olrn_{i \bs i} = \E[\ln(1-\delta+\delta B_i^\nat/B_j^\nat)].
\end{equation}
Likewise, $\olrn_{j \bs i}$ is computed by replacing $(E_j,C_{j \bs i})=(B_j,B_j/\delta)$ in an analogous way.
But note that $B_j$ and $B_j/\delta$ are linearly related for the lottery model, so their association is 
already symmetric, so
so $E_j^\nat=E_j$, $C_{j \bs i}^\nat = C_{j \bs i}$, and $\olrn_{j \bs i}=\olr_{j \bs i}$.
This type of reasoning is applied to the log-normal fecundities and beta fecundities versions of the 
lottery model, specifically, in the subsequent sections.

@@

\subsection{Theory and plotting simplifications for the beta fecundities lottery model}\label{SIsect:TheoryForLotteryBeta}

We first compute all terms of the decomposition (\ref{eq:decomp_overall_IGR}) of $\olr_{i \bs i}$ for the 
beta fecundities lottery model. Defining, for this section, $\beta_i=F_\beta^{-1} \circ \varphi(b_i)$ 
where $F_\beta$ is the CDF of a beta distribution with parameters $0.5,0.5$ and $\varphi$ is the CDF of 
a standard normal distribution, we have $B_i = \eta_i \beta_i$. Then, using the formulas of 
SI section \ref{SIsect:other_mechanisms}, we have
\begin{equation}
E_i = B_i = \eta_i \beta_i 
\end{equation}
\begin{equation}
B_j = \eta_j \beta_j 
\end{equation}
\begin{equation}
C_{i \bs i} = \frac{B_j}{\delta} = \frac{\eta_j}{\delta}\beta_j
\end{equation}
\begin{equation}
\overline{E}_i = \overline{B}_i = \frac{\eta_i}{2} 
\end{equation}
\begin{equation}
\overline{B}_j = \frac{\eta_j}{2} 
\end{equation}
\begin{equation}
\overline{C}_{i \bs i} = \frac{\eta_j}{2 \delta} 
\end{equation}
\begin{equation}
r_{i \bs i} = \ln\left(1-\delta+\delta \frac{B_i}{B_j} \right) = \ln\left(1-\delta+\delta\frac{\eta_i \beta_i}{\eta_j \beta_j}\right)
\end{equation}
\begin{equation}
\epsilon_i^0 = \ln\left( 1-\delta+\delta \frac{\eta_i}{\eta_j}  \right) 
\end{equation}
\begin{equation}
\olep_i^E = \E\left[ \ln\left( 1-\delta+\delta \frac{2\eta_i \beta_i}{\eta_j}  \right)  \right]-\epsilon_i^0
\end{equation}
\begin{equation}
\olep_i^C \E\left[ \ln\left(1-\delta+\delta\frac{\eta_i}{2\eta_j \beta_j}\right)  \right]-\epsilon_i^0
\end{equation}
\begin{equation}
\olep_i^{EC} = \E\left[ \ln\left( 1-\delta+\delta\frac{\eta_i \beta_i}{\eta_j \beta_j}  \right) \right]-\olep_i^E-\olep_i^C-\epsilon_i^0
\end{equation}
and
\begin{equation}
\olep_i^{(E\shp C)} = \E\left[ \ln\left( 1-\delta+\delta\frac{\eta_i \beta_i^\shp}{\eta_j \beta_j^\shp}  \right)  \right]-\olep_i^E-\olep_i^C-\epsilon_i^0,
\end{equation}
where $(\beta_i^\shp,\beta_j^\shp)=(F_\beta^{-1} \circ \varphi (b_i^\shp),F_\beta^{-1} \circ \varphi(b_j^\shp))$.
Here, $b_i^\shp$ and $b_j^\shp$ are standard normal and independent. Continuing,
\begin{equation}
\olep_i^{(EC)} = \E\left[ \ln\left( 1-\delta+\delta\frac{\eta_i \beta_i}{\eta_j \beta_j}  \right) \right]
-\E\left[ \ln\left( 1-\delta+\delta\frac{\eta_i\beta_i^\shp}{\eta_j \beta_j^\shp} \right) \right].
\end{equation}
Continuing,
\begin{equation}
\olep_i^{[EC]}=\E\left[ \ln\left( 1-\delta+\delta\frac{\eta_i \beta_i}{\eta_j \beta_j} \right) \right]-
\E\left[ \ln\left( 1-\delta+\delta\frac{\eta_i \beta_i^\nat}{\eta_j \beta_j^\nat}  \right)  \right],
\end{equation}
where $(\beta_i^\nat,\beta_j^\nat)=(F_\beta^{-1} \circ \varphi (b_i^\nat),F_\beta^{-1} \circ \varphi(b_j^\nat))$,
and $(b_i^\nat,b_j^\nat)$ has the same (standard normal) marginals and covariance as $(b_i,b_j)$ but is
bivariate normal. Note that a choice has been made here: we could have chosen $(b_i^\nat,b_j^\nat)$
to have the same (standard normal) marginals as $(b_i,b_j)$, but with covariance chosen so that
the Pearson, Spearman, or Kendall correlation of $E_i^\nat = \eta_i F_\beta^{-1} \circ \varphi (b_i^\nat)$
and $C_{i \bs i}^\nat = \eta_j F_\beta^{-1} \circ \varphi (b_j^\nat)/\delta$ is the same as that of 
$E_i = \eta_i F_\beta^{-1} \circ \varphi (b_i)$ and $C_{i \bs i} = \eta_j F_\beta^{-1} \circ \varphi (b_j)/\delta$.
But the choice we made seems the most natural under the circumstances.
Continuing, $\olep_i^{[E\nat C]}=\olep_i^{(E C)}-\olep_i^{[E C]}$ is available by subtraction, i.e.,
\begin{equation}
\olep_i^{[E\nat C]} = \E\left[ \ln\left(1-\delta+\delta\frac{\eta_i \beta_i^\nat}{\eta_j \beta_j^\nat}  \right)  \right] - 
\E\left[ \ln\left(1-\delta+\delta\frac{\eta_i \beta_i^\shp}{\eta_j \beta_j^\shp}  \right)  \right].
\end{equation}
This completes the calculation of all the terms of the decomposition 
(\ref{eq:decomp_overall_IGR}) of $\olr_{i \bs i}$ for the beta fecundities lottery model.

We next decompose $\olr_{j \bs i}$ for the beta fecundities lottery model [see (\ref{eq:decomp_overall_rjbsi})]. We have 
\begin{equation}
E_j = B_j = \eta_j \beta_j
\end{equation}
\begin{equation}
C_{j \bs i} = \frac{B_j}{\delta} = \frac{\eta_j \beta_j}{\delta}
\end{equation}
\begin{equation}
\overline{E}_j = \frac{\eta_j}{2}
\end{equation}
\begin{equation}
\overline{C}_{j \bs i} = \frac{\eta_j}{2\delta}
\end{equation}
\begin{equation}
\overline{B}_j = \frac{\eta_j}{2}
\end{equation}
\begin{align}
\olr_{j \bs i} &= \E\left[ \ln\left( 1-\delta+\frac{E_j}{C_{j \bs i}} \right) \right] \\
&= \E\left[ \ln\left( 1-\delta+\frac{B_j}{B_j} \right) \right] \\
&= 0 
\end{align}
\begin{equation}
\epsilon_j^0 = \ln \left( 1-\delta+\delta\frac{\eta_j /2}{\eta_j /2} \right) = 0
\end{equation}
\begin{align}
\olep_j^E &= \E\left[ \ln\left( 1-\delta+\delta\frac{2\eta_j \beta_j}{\eta_j} \right) \right] - \epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+2 \delta \beta_j \right) \right]
\end{align}
\begin{align}
\olep_j^C &= \E\left[ \ln\left( 1-\delta+\delta\frac{\eta_j}{2\eta_j\beta_j} \right) \right]-\epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+\frac{\delta}{2\beta_j} \right) \right]
\end{align}
\begin{align}
\olep_j^{EC} &= \E\left[  \ln\left( 1-\delta+\delta\frac{B_j}{B_j} \right)\right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= -\olep_j^E - \olep_j^C
\end{align}
\begin{align}
\olep_j^{(E\shp C)} &= \E\left[ \ln\left( 1-\delta+\delta\frac{B_j}{B_j^\shp} \right) \right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp} \right) \right]-\olep_j^E-\olep_j^C,
\end{align}
where $(\beta_j,\beta_j^\shp)=(F_\beta^{-1} \circ \varphi (b_j),F_\beta^{-1} \circ \varphi (b_j^\shp))$ and
$b_j^\shp$ is distributed in the same way as $b_j$ but is independent of it. Continuing,
\begin{align}
\olep_j^{(EC)} &= \E\left[ \ln\left(1-\delta+\delta \frac{B_j}{B_j} \right)  \right]-
\E\left[ \ln\left(1-\delta+\delta\frac{B_j}{B_j^\shp}  \right)  \right] \\
&= -\E\left[\ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp} \right)  \right].
\end{align}
Continuing,
\begin{align}
\olep_j^{[EC]} &= \E\left[ \ln\left(1-\delta+\delta \frac{B_j}{B_j} \right)  \right]-
\E\left[ \ln\left(1-\delta+\delta\frac{B_j^\nat}{B_j^\nat}  \right)  \right] \\
&= 0.
\end{align}
We showed already (SI section \ref{SIsect:TheoryForLotteryBoth} and 
\ref{SIsect:TheoryForLotteryLogNorm}) that this is $0$ because $E_j$ and $C_{j \bs i}$
are linearly related to each other. Continuing,
\begin{align}
\olep_j^{[E \nat C]} &= \olep_j^{(EC)}-\olep_j^{[EC]} \\
&= -\E\left[ \ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp}  \right)  \right].
\end{align}
This completes the calculation of all the terms of the decomposition of $\olr_{j \bs i}$
for the beta fecundities lottery model.

Using the above formulas, we observe that all the terms of the decomposition of $\olr_{i \bs i}$ depend on 
$\delta$, $\sigma$, and $\eta_i/\eta_j$, but not separately on $\eta_i$ or $\eta_j$. The terms of the decomposition
of $\olr_{j \bs i}$ depend on $\delta$ and $\sigma$, but do not depend on $\eta_i$ or $\eta_j$ at all.
As discussed in the main text and SI section \ref{SIsect:scaling_factors}, we consider two alternative
values for the scaling factors, $q_{ij}$, namely $q_{ij}=1$ and 
\begin{equation}
q_{ij} = \frac{\overline{B}_i}{\overline{B}_j (1-\delta) +\overline{B}_i \delta}.
\end{equation}
For the beta fecundities lottery model, this is
\begin{align}
q_{ij} &= \frac{\eta_i}{\eta_j(1-\delta)+\eta_i \delta} \\
&= \frac{\eta_i/\eta_j}{(1-\delta)+\frac{\eta_i}{\eta_j}\delta}.
\end{align}
Thus the $q_{ij}$ either do not depend on $\eta_i$ or $\eta_j$ at all or they depend on the quotient 
$\eta_i/\eta_j$, rather than depending separately on $\eta_i$ or $\eta_j$. Thus each of the terms of the 
decomposition (\ref{MT-eq:full_decomp}) from the main text
depends only on $\eta_i/\eta_j$ and not separately on $\eta_i$ or $\eta_j$, as stated in the Methods 
section of the main text.
